---
title: "old code"
author: "Kathryn Denning"
date: "2/27/2019"
output: html_document
---

```{r setup, include=FALSE}
###Plot 2: A heatmap showing the frequency of reports over days in the last month
Heatmap will show frequency of reports on the y-axis and time on the x-axis. I plan on using the viridis color palette.

```{r}
#cleaning date_time variable to be separate variables and making date interpretable using lubridate
plot2_time <- join1 %>% 
  separate(date_time, c("date", "time"), sep = " ") %>% 
  mutate(date = lubridate::mdy(date),
         state = state_abrv) 

#then getting frequncy of observations per state per day
plot2_state <- plot2_time %>% 
  group_by(state, date) %>% 
  count()

#doing the same but for region instead of state
##first need to merge datasets to have region information
join2 <- left_join(region_withfullstate, plot_freq_bystate_andregion2)

##now getting frequency
plot2_region_joined <-join2 %>% 
  group_by(region, date) %>% 
  count()

#heatmap of observations per state per day
ggplot(plot2 , aes(state, date)) +
    geom_tile(aes(fill = n)) +
scale_fill_viridis_c() +
  coord_flip()

#heatmap of observations per region per day
ggplot(plot2_region_joined, aes(region, date)) +
  geom_tile(aes(fill = n)) +
  scale_fill_viridis_c(name = "Frequency") +
  theme_minimal()
```

```{r}

```{r}
plot_freq_bystate_andregion %<>% 
  mutate(shape = as.factor(shape))

shape_plot_us <- plot_freq_bystate_andregion %>% 
  group_by(shape) %>% 
  filter(shape != "Unknown") %>% 
  count() %>% 
  arrange(-nn) %>% 
  na.omit() %>% 
  filter(nn > 10)

ggplot(shape_plot_us, aes(reorder(shape, -nn), nn, fill = shape)) +
  geom_col(position = "dodge") +
  #scale_fill_viridis_d() +
  labs(title = "Frequency of UFO observations by shape",
       y = "Frequency",
       x = "Shape") +
  theme_dark() + 
  theme(
    panel.background = element_rect(fill = "#2D2D2D"),
    legend.position = "none") +
  scale_fill_tron()
```
```
