---
title: "Final Project Visualizations - Draft 1"
author: "Kathryn Denning"
date: "2/25/2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

````{r setup and data import, echo = FALSE, warning = FALSE, message=FALSE, error=FALSE}
#install.packages("rio")
#install.packages("here")
#install.packages("tidyverse")
#install.packages("magrittr")
#install.packages("RColorBrewer")
#install.packages("ggridges")
#install.packages("XML")
#install.packages("janitor")
#install.packages("colorspace")
#install.packages("ggrepel")
library(rio)
library(here)
library(tidyverse)
library(magrittr)
library(RColorBrewer)
library(ggridges)
library(XML)
library(janitor)
library(colorspace)
#library(ggsci)
library(ggrepel)
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)

#read in data on all UFO citings in Jan 2019
x <- readHTMLTable("http://www.nuforc.org/webreports/ndxe201901.html")
#Turn it into a dataset
ufo_jan19 <- plyr::ldply(x, data.frame)
#power was out, so couldn't google how to add a key when there isn't one - so just wrote it into a csv and added in excel
write.csv(ufo_jan19,'ufo_jan19.csv')
ufo_jan19 <- read_csv("ufo_jan19.csv")

#cleaned dataset to be lowercase and to only include US states
clean <- ufo_jan19 %>%
  clean_names() %>% 
  filter(state != "AB") %>% 
  filter(state != "ON") %>% 
  filter(!is.na(state))

#Read in data of all ufo citings in Oregon ever
y <- readHTMLTable("http://www.nuforc.org/webreports/ndxlOR.html")
#Turn it into a dataset
ufo_oregon <- plyr::ldply(y, data.frame)
```

#Description of data:
The two main datafiles come from the National UFO Reporting Center (http://www.nuforc.org/index.html). One dataset includes all of the reported UFO sightings for January 2019. The other file contains all of the reported UFO sightings in Oregon on record. Both datasets include the following variables: date/time, city, state (though for the Oregon dataset this is all one state), shape (of the reported UFO), duration (of the object's appearance), summary (description by the reporter), and date posted (when it was posted on the website). A .id column appeared when converting both datafiles into a dataframe, but is not on the website and all datafilesand will be deleted.

I also used data on the state population size from the US Census (https://www.census.gov/newsroom/press-kits/2018/pop-estimates-national-state.html),county populaiton size from an education resource called "Together we teach" (http://www.togetherweteach.com/TWTIC/uscityinfo/37or/37or.html), and city/county association from http://www.ehealthlink.com/citytocounty.htm.

#Description of plots

All plots will be made for a general public audience, though the scientific(?) community interested in UFO sightings might also be interested.

###Plot 1A: Choropleth of frequency of reports across US by location
Plot showing the shape of the country with colors changing on the map according to the density of the reports, using changes in color like a heatmap and ideally the viridis color palette. I plan on doing this at the city level within each state, but hopefully can still have the outline of each state visible in the map of the US as well.

```{r plot 1 - chloropleth version 1}
state <- ggplot2::map_data("state")

#need to have a variable with the full state names in lowercase to merge datasets
state_fullname <- datasets::state.name #calling dataset with state names
state_fullname <- sapply(state.name, tolower) #making them lowercase

state_abbrv <- datasets::state.abb #calling dataset with abbreviations to use as key when merging dataset to get full names in main dataset so that it can be combined with dataset used to make map of US

state_key <- data.frame(
  state_full = state_fullname,
  state_abrv = state_abbrv
) #making dataset with the two character strings state_fullname and state_abbrv

state_key %<>% clean_names

clean %<>% 
  mutate(state_abrv = state)

join1 <- left_join(clean, state_key) #joining dataset with fullnames to cleaned ufo dataset

#getting frequency of UFO citings per state
plot1 <- join1 %>% 
  group_by(state_full) %>% 
  count()

#cleaning dataset that has state information for plot of US so that it can be merged
state_fixed <- state %>% 
  mutate(state_full = region) %>% 
  select(-region) %>% 
  clean_names

#merging data with frequency of UFO with plot dataset
plot1_freq_merged <- left_join(state_fixed, plot1)

#code that removes axes from plot of US
ditch_the_axes <- theme(
  axis.text = element_blank(),
  axis.line = element_blank(),
  axis.ticks = element_blank(),
  panel.border = element_blank(),
  panel.grid = element_blank(),
  axis.title = element_blank()
  )

#code that makes US base
state_base <- ggplot(data = plot1_freq_merged, 
                     mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
  geom_polygon(color = "white", fill = "white")

#code that fills the states with frequency data
state_filled <- state_base + 
  geom_polygon(data = plot1_freq_merged, aes(fill = n), color = "white") +
      geom_polygon(color = "white", fill = NA) +
      scale_fill_continuous_sequential("viridis", name = "Frequency") +
  ditch_the_axes

#adding title
state_filled + 
  labs(title = "Frequency of UFO citings per state",
       caption = "Data from National UFO Reporting Center and US Census Bureau")
```

After making the first chloropleth, I decided the patterns it is showing might be a little misguided, as the frequency of observations seems to be confounded with the population size of the state. I decided to re-do the plot to represent the proportion of observations according to the state population.

###Plot 1B: Chloropleth of proportion

```{r plot 1 - chloropleth with proportion of observations per population}
#importing dataset with population stats from the United States Census Bureau
pop <- read_csv("nst-est2018-alldata.csv")

#cleaning the formatting of the names
pop %<>% 
  clean_names() 

#changing the state name variable so that it will be the same when merging later and making data lowercase
pop$state_full <- tolower(pop$name)

#cleaning dataset to exclude areas that weren't in my UFO dataset
pop %<>%
  filter(state_full != "puerto rico") %>% 
  filter(state_full != "district of columbia") %>% 
  filter(state_full != "united states")

#making dataset that is only the 50 states, ignoring the four regions for the moment
pop_state <- pop %>%
  filter(state_full != "northeast region") %>% 
  filter(state_full != "midwest region") %>% 
  filter(state_full != "south region") %>% 
  filter(state_full != "west region")

#merging population dataset with the data on the frequency of UFO observations
pop_state_freq_merged <- left_join(pop_state, plot1_freq_merged)

#converting the frequency of observations to be a proportion
freq_bypop_state <- pop_state_freq_merged %>% 
  group_by(state_full) %>% 
  mutate(freq_bypop = ((n/popestimate2018))) %>% 
  summarize(freq_bypop_perstate = mean(freq_bypop))

#adding the proportion back into the main dataset
pop_state_freq_merged1 <- left_join(plot1_freq_merged, freq_bypop_state)

#base of US for the plot
state_base2 <- ggplot(data = pop_state_freq_merged1 , mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
  geom_polygon(color = "white", fill = "white")

#filling in the states with the proportion data
state_filled2 <- state_base + 
  geom_polygon(data = pop_state_freq_merged1 , aes(fill = freq_bypop_perstate), color = "white") +
      geom_polygon(color = "white", fill = NA) +
      scale_fill_continuous_sequential("viridis", name = "Proportion") +
  ditch_the_axes

#adding labels
state_filled2 + 
  labs(title = "Proportion of UFO citings per state",
       subtitle = "Proportion determined by number of citings divided by population size",
       caption = "Data from National UFO Reporting Center and US Census Bureau")

# ASH: I agree with your decision to control for a state's population, and the graph itself looks fantastic. I actually love the default viridis color palette! For the final product, it may be cool to combine plot 1a and 1b via cowplot or something similar. Also, with plot 1b, I'd try converting the scale into something a bit easier to interpret. Personally, I try to avoid scientific notation.

```

I then decided that another way to visualized this information might be to make a barplot showing the frequency of observations by state and region with state organized in descending order in regards to population, to show outlier states with more observations that other states around their same size in population.

###Plot2: Bar plot showing frequency of observations per state with state in descending order by population

```{r plot2}
#made the population dataset to include a variable of region
pop_region <- pop %>%
  select(c(state_full, popestimate2018, region)) %>%  
  filter(state_full != "northeast region") %>% 
  filter(state_full != "midwest region") %>% 
  filter(state_full != "south region") %>% 
  filter(state_full != "west region") %>% 
  mutate(region = factor(region, labels = c("Northeast",
                                              "Midwest",
                                              "South",
                                              "West"))) 


#merged population region dataset with overall dataset with dataset with state names and regions to create key that can merge with the UFO obersevation dataset
region_withfullstate <- left_join(state_key, pop_region)

#merged region data with UFO observation data
pop_freq_state_region <- left_join(region_withfullstate, clean)

#Make a frequency of observations from this dataset
freq_bystate <- pop_freq_state_region %>% 
  group_by(state_abrv) %>% 
  count()

#merged the frequency variable back into the overall dataset
plot_freq_bystate_andregion <- left_join(pop_freq_state_region, freq_bystate)

#got rid of variables I didn't want and repeat rows 
plot_freq_bystate_andregion2 <- plot_freq_bystate_andregion %>% 
  select(state_abrv, popestimate2018, region, n) %>% 
  unique()

#bar plot
ggplot(plot_freq_bystate_andregion2, aes(reorder(state_abrv, popestimate2018), n)) +
  geom_col(aes(fill = region)) +
  ylim(0,25) +
  coord_flip() +
  theme_minimal() +
  scale_fill_viridis_d(name = "Region") +
  labs(title = "Frequency of UFO observations by state and region",
       subtitle = "States in descending order by population size",
       y = "Frequency of UFO observations",
       x = "State",
       caption = "Data from National UFO Reporting Center and US Census Bureau")

# ASH: I understand/appreciate you trying to see if this method is better at depicting the message gathered from your first plot. But, with the addition of coloring by region, I think it would be better to reduce cognitive load/change the focus of this plot by reordering states by n rather than their population. That way we can better observe any potential patterns as a function of region. I'd also consider increasing the width of your bars or the height of the figure itself (your state labels are overlapping).

```

Started with this heatmap, did it by state and by region, and realized that, without observations each day, a heatmap really wasn't going to work too well and there wasn't enough data overtime (only a month) to really bin larger time intervals. If you have suggestions of how to improve this, please let me know.

###Plot 3A: Choropleth of frequency of reports across Oregon by location
Plot showing the shape of the state of Oregon with colors changing on the map according to the density of the reports, using changes in color like a heatmap and ideally the viridis color palette. I would like to have either county outlines on the state and cities also identified on the map for reference (like Eugene, Portland, Bend, McMinville (UFO festival location, has to be identified!)).

```{r chloropleth of frequency in oregon ever}
#cleaning the oregon dataset
ufo_oregon %<>% clean_names() 

#importing dataset with information about city, county, and population information
oregon_city_county <- import("oregon_city_county.csv")

#joining dataset of ufo data with oregon city and county data
oregon_join1 <- left_join(oregon_city_county, ufo_oregon)

#cleaning new dataset
oregon_join1 %<>% 
  na.omit() %>% 
  select(-id)

#getting informtion on the lat and long of each county for plotting
county <- ggplot2::map_data("county")

#changing the variables in county to have a key that will match ufo_oregon dataset
oregon_county <- county %>% 
  filter(region == "oregon") %>% 
  mutate(county = subregion)

oregon_join1$county <- tolower(oregon_join1$county)

#merging the county lat long dataset with the oregon_join1 dataset
oregon_join2 <- left_join(oregon_join1, oregon_county)

#getting the frequency of observations per county
oregon_plot_freq <- oregon_join2 %>% 
  group_by(county) %>% 
  count()

#joining the frequency back with the main dataset
oregon_join3 <- left_join(oregon_join2, oregon_plot_freq)

#dropping region and subregion from new overall dataset
oregon_join3 %<>% 
  select(-c(region, subregion))

#creating the dataset for the state base for the plot
oregon_base <- state_fixed %>% 
  filter(state_full == "oregon")

#making the base of oregon for the plot
oregon_base <- ggplot(data = oregon_base, 
                     mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
  geom_polygon(color = "white", fill = "white")

#code that fills oregon with frequency data
oregon_filled <- oregon_base + 
  geom_polygon(data = oregon_join3, aes(fill = n), color = "white") +
      geom_polygon(color = "white", fill = NA) +
      scale_fill_continuous_sequential("viridis", name = "Frequency") +
  ditch_the_axes

#adding title
plot_title_freq <- oregon_filled + 
  labs(title = "Frequency of UFO citings per county in Oregon",
       subtitle = "Top five counties labeled",
       caption = "Data from National UFO Reporting Center and US Census Bureau")

#figuring out top 5
oregon_join3 %>% 
  select(county, n, group) %>% 
  arrange(-n) %>% 
  unique() %>% 
  filter(n > 6693)

#Adding labels for the top 5 counties
plot_title_freq + theme(panel.background = element_rect(fill = "gray22")) +
  annotate("text", 
           label = "Multnomah", 
           y = 45.5, 
           x = -122.2, 
           color = "white", 
           size = 2.5,
           fontface = "bold") +
  annotate("text", 
           label = "Lane", 
           x = -122.7, 
           y = 44, 
           color = "white", 
           size = 2.5,
           fontface = "bold") +
    annotate("text", 
           label = "Washington", 
           x = -123.3, 
           y = 45.5, 
           color = "white", 
           size = 2.5,
           fontface = "bold") +
    annotate("text", 
           label = "Marion", 
           x = -122.1, 
           y = 44.8, 
           color = "white", 
           size = 2.5,
           fontface = "bold") +
   annotate("text", 
           label = "Clackamas", 
           x = -122.2, 
           y = 45.2, 
           color = "white", 
           size = 2.5,
           fontface = "bold")

# ASH: Oooh, this is very pretty! One way to enhance this plot could be by printing the text in a different color? The white is a litte hard to read when it overlaps county lines.

```

###Plot 3B: Choropleth of proportion of reports across Oregon by location

```{r chloropleth of proportion in oregon ever}
#finding the proportion of observations per county
oregon_plot_proportion <- oregon_join3 %>% 
  group_by(county) %>% 
  mutate(freq_bypop = (n/county_pop)) %>% # ASH: To be clear, here you are controlling for population size for a given county?
  summarize(freq_bypop_percounty = mean(freq_bypop))

#adding proportion back into overall dataset
oregon_join4 <- left_join(oregon_join3, oregon_plot_proportion)

#filling in base with county proportion information
oregon_filled_pro <- oregon_base + 
  geom_polygon(data = oregon_join4, aes(fill = freq_bypop_percounty), color = "white") +
      geom_polygon(color = "white", fill = NA) +
      scale_fill_continuous_sequential("viridis", name = "Proportion") +
  ditch_the_axes

#adding title
plot_title <- oregon_filled_pro + 
  labs(title = "Proportion of UFO citings per county in Oregon",
       subtitle = "Top five counties labeled",
       caption = "Data from National UFO Reporting Center and US Census Bureau") 

#Figuring out top 5 counties
county_names <- oregon_join4 %>% 
  select(county, freq_bypop_percounty, group) %>% 
  arrange(-freq_bypop_percounty) %>% 
  unique() %>% 
  filter(freq_bypop_percounty > 0.06523439)

#Adding labels for the top 5 counties
plot_title + 
  theme(panel.background = element_rect(fill = "gray22")) +
  annotate("text", 
           label = "Baker", 
           y = 44.7, 
           x = -117.7, 
           color = "white", 
           size = 3.5,
           fontface = "bold") +
  annotate("text", 
           label = "Lane", 
           x = -122.7, 
           y = 44, 
           color = "white", 
           size = 3.5,
           fontface = "bold") +
    annotate("text", 
           label = "Curry", 
           x = -124.5, 
           y = 42.4, 
           color = "white", 
           size = 3,
           fontface = "bold") +
    annotate("text", 
           label = "Tillamook", 
           x = -124.2, 
           y = 46, 
           color = "white", 
           size = 3,
           fontface = "bold") +
   annotate("text", 
           label = "Lincoln", 
           x = -124.2, 
           y = 44.7, 
           color = "white", 
           size = 3,
           fontface = "bold") 

# ASH: See previous comment about text color. Perhaps the white could work if you move the text around a bit (e.g., shift Tillamook, Lincoln, and Curry to the left).

```

### Ashley's peer review breakdown:

**Three areas of strength:**

  + Excellent use of commentary; I knew what you were trying to achieve with each line of code!
  + Super creative/clever idea with exceptional execution:
     + The fact that you created so many geospatial maps is incredible! Very impressive! I have yet to try this myself because it seems daunting. You also used functions not covered in class (e.g., sapply—line 75), which is quite sophisticated in my opinion. 
  + Great graphical presentation:
      + Good use of color-blind friendly palette

**At least one thing I learned:**

  + In the past, I had to do a lot of googling/examining old assignments to find another dataset that I could join to an existing dataset (I needed info not in the existing file). I never thought of creating my own key then importing it like you did yourself. 
  + I also didn't know that you could call the viridis color palette via scale_fill_continuous_sequential--good to know! I always use scale_fill_viridis_c()
